Demand Forecasting of individual Probability Density Functions with Machine Learning

### 2020-09-06 LaTex/paper.tex
forgot accuracy formula, some renaming

Many statistical and most machine learning methods do not provide a full probability density distribution as result. Instead, these methods typically predict a single numerical quantity that is then compared to the observed concrete realization of the random variable using metrics such as the mean squared error (MSE), mean absolute deviation (MAD) or others. In the setting of a retailer, the observed quantity is the sales of individual products (denoted by $y$) and most machine learning approaches would then predict a single number $\hat{y}$ as a direct estimate of the sales. However, reducing the prediction to a single number does not allow to account for the uncertainty of the prediction or the dynamics of the system. Instead, it is imperative to predict the full PDF for each prediction to be able to optimize the subsequent operational decision. Unfortunately, most statistical or machine learning methods that predict full individual probability functions lack quantitative or at least qualitative evaluation methods to assess whether the full distribution has been forecast correctly, in particular in the tails of the distribution.

For an estimation of the determining parameters of an assumed functional form for the PDF, assessing the correctness of the PDF model output refers to the evaluation of the accuracy of the prediction of the different determining parameters. In the case of the negative binomial distribution used in this work, we have to verify that mean and variance are determined accurately, as well as checking that the choice of the underlying model can describe the observed data.

In the following, we will show how different visualizations of the observed cumulative distribution function (CDF) values can be used to evaluate the quality of the predicted PDFs. Although we limit the following discussion to the negative binomial model, the method can be applied generally to any representation of a probability density distribution, even if the PDF is obtained empirically.

@@

For a first qualitative assessment, we make use of the probability integral transform, see e.g. \cite{Angus1994,casella2002statistical}, which states that a random variable distributed according to the CDF of another random variable is uniformly distributed. We therefore expect that distribution of the corresponding actually observed CDF values of the individual PDF predictions is uniform if the predicted PDF is calibrated correctly, {\em regardless} of the shape of the predicted distribution. Any deviation can be interpreted as a hint that the predicted PDF is not fully correct \cite{diebold1998vevaluating}.

@@

If the CDF is continuous and strictly increasing, then the inverse of the CDF, $F^{-1}(y)$, exists and is a unique real-valued number $x$ for each $y \in [0,1]$, so that we can write $F(x) = y$. The latter defines the inverse quantile function, because we can define the quantile $\tau$ of the probability distribution $f(x)$ as:

@@

In case of discrete probability functions, such as the negative binomial function, the same argument still holds, but the definition of the inverse quantile function is replaced by the generalized inverse: $F^{-1}(y) = \mathrm{\inf \left \{x : F(x)>y\right  \} }$ for $y \in [0,1]$, see e.g. \cite[p. 54]{casella2002statistical}. In order to obtain a uniform distribution for discrete PDFs that is comparable to the case of continuous distributions, the histogram holding the values of the CDF is filled using random numbers according to the intervals of the CDF. For example, if the sales of zero items accounts for 75 percent of the observed sales distribution for this item, the value of the CDF function that is used to fill the histogram in case of zero actual sales is randomly chosen in the interval $[0, 0.75]$. Proceeding similarly for all other observed values, with the intervals from which to randomly choose values to fill in the histogram defined by the CDF values of the corresponding discrete sales value and the one below (e.g. for 3 actual sales: random pick between discrete CDF values for 2 and 3), the resulting histogram of CDF values is again uniform as in the case of a continuous PDF.

A histogram of the observed quantiles (or CDF values) for each individual PDF prediction is therefore expected to be uniformly distributed in $[0,1]$, if the predicted probability distribution $f(x)$ is correctly calibrated. This is illustrated in \fig \ref{fig:cdf_histos}, which shows the distribution of observed quantiles for five different cases. If both the choice of the model and the model parameters are estimated correctly, we would expect the uniform distribution. If the mean or the variance are not estimated correctly, the resulting distribution will show a distinct deviation from this uniform behavior.

\begin{figure}
\begin{center}
\includegraphics[width=2.5cm]{../figures/cdf_truth_uniform}
\includegraphics[width=2.5cm]{../figures/cdf_truth_broad}
\includegraphics[width=2.5cm]{../figures/cdf_truth_narrow}
\includegraphics[width=2.5cm]{../figures/cdf_truth_over}
\includegraphics[width=2.5cm]{../figures/cdf_truth_under}
\caption{\label{fig:cdf_histos} Quantile distribution for different cases of estimating the full probability distribution compared to the expected distribution:
correct prediction (``uniform''), variance overestimated (``broad''), variance underestimated (``narrow''), mean overestimated (``over''), mean underestimated (``under'')}
\end{center}
\end{figure}

\subsubsection{Quantile Profile Plot}

We now refine the method from sec. \ref{sec:cdf_histo} by comparing (in the sense of higher or lower) the quantiles of the observed events, i.e., the sales records, with specified quantile values. In order to do so, we start again from the predicted values for the mean and variance from which a negative binomial distribution is constructed for each prediction, for example, for the predicted demand for a single product sold on a single day in a given sales location. Each of these predicted negative binomial PDFs is then transformed to its CDF. Note that for simplicity, we always refer to the negative binomial model in this description, however, the general approach is valid for any PDF.

Then we compare the actual observed sales value (corrected for censored data if necessary) to different quantiles of the corresponding predicted distribution for each data record and average over a larger data sample. For example, if we wanted to check that the median of the distribution, corresponding to the $0.5$ quantile, is predicted correctly by the machine learning model, we would compare the quantile value $0.5$ to the ratio of quantile values (again randomly chosen from the corresponding range of CDF values for discrete target values, as described in sec. \ref{sec:cdf_histo}) of observed sales records being lower/higher than $0.5$. In other words, in case of the median, 50 percent of the ex post observed target values should be observed below the median of the corresponding individual predicted PDF and 50 percent above.

In order to judge whether the overall shape of the predicted distributions is predicted correctly, we repeat this procedure for a range of quantiles, for example $q = 0.1, 0.2, \ldots, 0.9$. However, we are free to choose which quantiles to look at, and in specific situations it might be advisable to look at the tails of the distribution in more detail, to make sure that even relatively rare events are estimated correctly by the machine learning algorithm, and add more quantiles for comparison in the region between, say $q = 0.95$ and $q = 0.99$. In the following, we call this method {\em quantile profile plot}. Profile plots are akin to scatter plots and described in more detail in appendix \ref{sec:profile}.

\fig \ref{fig:invquant_example} illustrates five different collections of quantile profile plots (each collection comparing to 6 specified quantile values, namely $q = 0.1$, $q = 0.3$, $q = 0.5$, $q = 0.7$, $q = 0.9$, and $q = 0.97$), for separate sets of exemplary PDF estimations and observed data combinations. The dashed horizontal lines indicates the fraction we expect, i.e., the specified quantile value, if the predictions are correct. For example, for the median, the line at $q = 0.5$ indicates that  50 percent of all PDF prediction and observed data combinations in a given data set should fall above the line, and 50 percent should fall below the line. The observation of the number of samples, indicated with different markers, that do in fact fall above and below a particular line, then allows the evaluation of the accuracy of PDF estimations. In case that the PDFs are not estimated correctly, the fractions will deviate from their expected values and the corresponding profile plot allows to judge whether for example the tails of the predicted distribution describing rare events are particularly problematic.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/invquant_example}
\caption{\label{fig:invquant_example} Profile plots of quantiles comparing all predicted PDFs to the corresponding observed events. In the leftmost two columns, the outcome is illustrated if the estimate of the variance is biased, the center and center-right columns illustrate cases for which the mean estimation is biased, and the far right column shows the expected behavior if all predictions are correct.}
\end{center}
\end{figure}

In the same way as the method of filling all observed CDF values of the individual PDF predictions in a histogram and compare to a uniform distribution (described in sec. \ref{sec:cdf_histo}), quantile profile plots do not work on individual PDF predictions but require a statistical population. However, calculating the fractions of the quantile profile plots globally, i.e., over all samples in the data set, might not reveal certain deficiencies for a subset, e.g., a specific store or product group in the example of retail demand forecasts. Therefore, we combine the approach described above with the method of profile plots, where the quantity on the $x$-axis of the quantile profile plot can be any variable of the data set at hand.

In summary, this approach has two major benefits compared to the method discussed in sec. \ref{sec:cdf_histo}: First, by explicitly visualizing several different quantiles, the quantile profile plot reveals which part of the predicted PDF, such as the tails of the probability distribution, are particularly problematic. Second, by showing the dependency from the (arbitrary) variable on the $x$-axis, the quantile profile plot reveals deviations of the predicted PDF from the actuals for different parts, e.g. specific categories, of that variable. Two examples for this can be found in Figs. \ref{fig:invquant_mean} and \ref{fig:invquant_dayofweek} in the next section.

@@

The methods discussed  so far allow a detailed qualitative evaluation of PDF predictions. However, in order to also quantify the quality of the PDF predictions, a metric is required that assesses the difference between the distributions. Therefore, we want to compare the CDF histogram of the predicted PDFs with the expected  uniform distribution and define a metric in the range between 0 and 1, such that the metric takes the value of 1 when both distributions agree perfectly. Several approaches that measure the difference between two probability distributions are suggested in the literature such as  the first Wasserstein distance \cite{olkin1982}, the Kullback-Leibler divergence \cite{kullback1951}, and the Jensen-Shannon divergence  \cite{dagan1997}.

The first Wasserstein distance, also known as earth mover distance (EMD), represents a distance  defined between two probability distributions on a given metric space, and for our purposes here can defined by:
\begin{equation}
\text{EMD}(P, Q) = 2 \cdot \frac{\sum_{k=1}^N |F_P(x_k) - F_Q(x_k)|}{N},
\end{equation}
where $F_P(X)$ and $F_Q(X)$ are the CDFs of the two PDFs $P(X)$ and $Q(X)$, respectively, and $x_k$ denotes the average value of $X$ in bin $k$, with $X$ being divided in $N$ bins. Compared to the other metrics given below, the first Wasserstein or earth mover distance has the additional benefit, that it takes a specific metric space into account, meaning that it depends on the distance of potential deviations.

The Kullback-Leibler divergence is a measure of how one probability distribution diverges from a second expected probability distribution. For PDFs $P(X)$ and $Q(X)$, again with $X$ being divided in $N$ bins $k$, the Kullback-Leibler divergence from Q to P is defined as:
\begin{equation}
D_{\text{KL}}(P \parallel Q) = \sum _{k=1}^N P(x_k) \log \left({\frac{P(x_k)}{Q(x_k)}}\right),
\end{equation}
where the logarithm can be either base-2 or the natural logarithm.

The Jensen-Shannon divergence (JSD) can be seen as a symmetrized and smoothed version of the Kullback-Leibler divergence:
\begin{equation}
D_{\text{JSD}}(P \parallel Q) = 0.5  \cdot (D_{\text{KL}}(P \parallel M) + D_{\text{KL}}(P \parallel M)),
\end{equation}
where $M = 0.5  \cdot (P + Q)$.

@@

\noindent
In this model focusing on the variance, we use the same set of features as for the mean model described above in sec. \ref{sec:example_mean}, except for dropping most of the two-dimensional combinations (only keeping \texttt{item\_id} - \texttt{store\_id}, \texttt{store\_id} - day of week, and \texttt{item\_id} - event type features) and adding the in-sample mean prediction as feature. An example for the resulting PDF and CDF predictions for item \texttt{FOODS\_3\_516} in store \texttt{TX\_3} on 2016-05-06 is shown in \fig \ref{fig:pdf_example}.

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/pdf}
\includegraphics[width=4cm]{../figures/cdf}
\caption{\label{fig:pdf_example} Predicted PDF (left) and CDF (right) for a specific product-location-date combination.}
\end{center}
\end{figure}

@@

Using the profile plots introduced earlier, the prediction quality of the full predicted PDF can be assessed in more detail. In \fig \ref{fig:invquant_dayofweek}, each column corresponds to a day of the week, using 0 for Mondays, 1 for Tuesdays, etc., and the considered quantile values are 0.1, 0.3, 0.5, 0.7, 0.9, and 0.97. Again, we compare the outcome of our negative binomial model to a simpler Poisson assumption. Aggregating over all stores, items, and sales dates (independently for each day of the week), we can see that for the negative binomial model each quantile is predicted relatively well, except for Sundays showing a tendency for overprediction of the mean (see \fig \ref{fig:invquant_example}). The Poisson model, on the other hand, shows a significant pattern of too narrow PDF predictions across all days of the week.

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/invquant_dayofweek_nbinom}
\includegraphics[width=4cm]{../figures/invquant_dayofweek_poisson}
\caption{\label{fig:invquant_dayofweek} Quantile profile plot for the different days of week on the $x$-axis (from Monday to Sunday) aggregated over all product-location-day combinations in the test period, for our negative binomial model (left) and a simpler Poisson model for comparison (right).}
\end{center}
\end{figure}

\fig \ref{fig:invquant_mean} shows the quantile profile plot using the predicted mean as the $x$-axis of the graph, meaning that the distributions are grouped such that the 15 columns correspond to mean predictions in 12 intervals with widths of 5 from 0.0 to 60.0 and 3 remaining intervals $[60.0, 70.0]$, $(70.0, 80.0]$, and $(80.0, 100.0]$ (with mean predictions higher than 100.0 included in the highest interval), while we aggregate over all locations, items, and sales dates. The  quantile values are chosen to be the same as in \fig \ref{fig:invquant_dayofweek}. While the Poisson model again shows a significant pattern of too narrow PDF predictions across the full range of mean prediction values, we can also see several deviations from the expected uniform behavior for the negative binomial model. For mean predictions around $30$, the negative binomial model deviates significantly from the expected behavior, with the shape of the deviations pointing towards too broad PDF predictions (as also observed in the CDF histogram above). And for mean predictions higher than $80$, the pattern of the deviations indicate an underprediction of the mean parameter. The deviations at very low predicted mean values reflect the complications of slow-sellers described in the context of CDF histograms above.

@@

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/invquant_yhat_mean_nbinom}
\includegraphics[width=4cm]{../figures/invquant_yhat_mean_poisson}
\caption{\label{fig:invquant_mean} Quantile profile plot for mean predictions on the $x$-axis aggregated over all product-location-day combinations in the test period, for our negative binomial model (left) and a simpler Poisson model for comparison (right).}
\end{center}
\end{figure}

The quantitative results for the CDF accuracy of our PDF predictions for different metrics (Wasserstein distance as well as Kullback-Leibler and Jensen-Shannon divergence with base-2 logarithm) described in sec. \ref{sec:cdf_acc}, calculated over all product-location-day combinations in the test period, can be found in \tab \ref{tab:cdf_acc}. As benchmark, we again compare against a Poisson model assumption using the same mean predictions. As expected from the qualitative findings in \fig \ref{fig:cdf_demand}, the negative binomial PDF predictions show a significant improvement over the simpler Poisson model.

\begin{table}[h!]
\begin{center}
\caption{CDF accuracy for negative binomial and Poisson PDF predictions for different metrics calculated over all product-location-day combinations in the test period.}
\label{tab:cdf_acc}
\begin{tabular}{c|c|c}
 & \textbf{NBD} & \textbf{Poisson} \\
\hline
\textbf{EMD} & 0.967 & 0.850 \\
\textbf{KL\_2} & 0.994 & 0.933 \\
\textbf{JSD\_2} & 0.998 & 0.981
\end{tabular}
\end{center}
\end{table}

@@

Once the full probability density functions are forecast, their evaluation poses significant challenges:
Common metrics, such as the mean absolute deviation or mean squared error, only take point estimates into account and are generally not suitable for the evaluation of predicted probability distributions. Furthermore, since the distributions are predicted for individual events, statistical techniques aiming at the comparison of distributions using many events cannot be used in general. Instead, novel techniques using both the histogram of the probability integral transformation as well as quantile profile histograms allow a detailed investigation into the behavior of the predictions. Finally, a quantitative assessment resulting in a single number can be obtained using metrics such as the earth mover distance in a comparison between the histogram of the probability integral transformation and the expected uniform distribution.

>>>>

Many statistical and most machine learning methods do not provide a full probability density distribution as result. Instead, these methods typically predict a single numerical quantity (denoted by $\hat{y}$) that is then compared to the observed concrete realization of the random variable (denoted by $y$) using metrics such as the mean squared error (MSE), mean absolute deviation (MAD) or others. In the setting of a retailer, the observed quantity is the sales of individual products and most machine learning approaches would then predict a single number as a direct estimate of the sales. However, reducing the prediction to a single number does not allow to account for the uncertainty of the prediction or the dynamics of the system. Instead, it is imperative to predict the full PDF for each prediction to be able to optimize the subsequent operational decision. Unfortunately, most statistical or machine learning methods that predict full individual probability functions lack quantitative or at least qualitative evaluation methods to assess whether the full distribution has been forecast correctly, in particular in the tails of the distribution.

For an estimation of the determining parameters of an assumed functional form for the PDF, assessing the correctness of the PDF model output refers to the evaluation of the accuracy of the prediction of the different parameters. In the case of the negative binomial distribution used in this work, we have to verify that mean and variance are determined accurately, as well as checking that the choice of the underlying model can describe the observed data.

In the following, we will show how different visualizations of the observed cumulative distribution function (CDF) values can be used to evaluate the quality of the predicted PDFs. Although we limit the following discussion to the negative binomial model, the method can be applied generally to any representation of a probability density distribution, even if it is obtained empirically.

@@

For a first qualitative assessment, we make use of the probability integral transform, see e.g. \cite{Angus1994,casella2002statistical}, which states that a random variable distributed according to the CDF of another random variable is uniformly distributed. We therefore expect that the distribution of the actually observed CDF values of the corresponding individual PDF predictions is uniform, if the predicted PDF is calibrated correctly, {\em regardless} of the shape of the predicted distribution. Any deviation can be interpreted as a hint that the predicted PDF is not fully correct \cite{diebold1998vevaluating}.

@@

If the CDF is continuous and strictly increasing, then the inverse of the CDF, $F^{-1}(y)$, exists and is a unique real-valued number $x$ for each $y \in [0,1]$, so that we can write $F(x) = y$. The inverse of the CDF is also called the quantile function, because we can define the quantile $\tau$ of the probability distribution $f(x)$ as:

@@

In case of discrete probability functions, such as the negative binomial function, the same argument still holds, but the definition of the quantile function is replaced by the generalized inverse: $F^{-1}(y) = \mathrm{\inf \left \{x : F(x)>y\right  \} }$ for $y \in [0,1]$, see e.g. \cite[p. 54]{casella2002statistical}. In order to obtain a uniform distribution for discrete PDFs that is comparable to the case of continuous distributions, the histogram holding the values of the CDF is filled using random numbers according to the intervals of the CDF. For example, if the sales of zero items accounts for 75 percent of the observed sales distribution for this item, the value of the CDF function that is used to fill the histogram in case of zero actual sales is randomly chosen in the interval $[0, 0.75]$. Proceeding similarly for all other observed values, with the intervals from which to randomly choose values to fill in the histogram defined by the CDF values of the corresponding discrete sales value and the one below (e.g. for 3 actual sales: random pick between discrete CDF values for 2 and 3), the resulting histogram of CDF values is again uniform, as in the case of a continuous PDF.

A histogram of the actually observed CDF values for each individual PDF prediction (see \fig \ref{fig:pdf_example} for an example) is therefore expected to be uniformly distributed in $[0,1]$, if the predicted probability distribution $f(x)$ is correctly calibrated. This is illustrated in \fig \ref{fig:cdf_histos}, which shows the distribution of observed CDF values for five different cases. If both the choice of the model and the model parameters are estimated correctly, we would expect the uniform distribution. If the mean or the variance are not estimated correctly, the resulting distribution will show a distinct deviation from this uniform behavior.

\begin{figure}
\begin{center}
\includegraphics[width=2.5cm]{../figures/cdf_truth_uniform}
\includegraphics[width=2.5cm]{../figures/cdf_truth_broad}
\includegraphics[width=2.5cm]{../figures/cdf_truth_narrow}
\includegraphics[width=2.5cm]{../figures/cdf_truth_over}
\includegraphics[width=2.5cm]{../figures/cdf_truth_under}
\caption{\label{fig:cdf_histos} Histogram of actually observed CDF values for different cases of estimating the full PDF compared to the expected uniform distribution:
correct prediction (``uniform''), variance overestimated (``broad''), variance underestimated (``narrow''), mean overestimated (``over''), mean underestimated (``under'')}
\end{center}
\end{figure}

\noindent
It should be noted that the method requires a sufficient sample size, as a too low number of observations of course leads to a discretization bias.

\subsubsection{Inverse Quantile Profile Plot}

We now refine the method from sec. \ref{sec:cdf_histo} by comparing (in the sense of higher or lower) the CDF values of the observed events, i.e., the sales records, with specified quantiles. In order to do so, we start again from the predicted values for the mean and variance from which a negative binomial distribution is constructed for each prediction, for example, for the predicted demand for a single product sold on a single day in a given sales location. Each of these predicted negative binomial PDFs is then transformed to its CDF. Note that for simplicity, we always refer to the negative binomial model in this description, however, the general approach is valid for any PDF.

Then we compare the actual observed sales value (corrected for censored data if necessary) to different quantiles of the corresponding predicted distribution for each data record and average over a larger data sample. For example, if we wanted to check that the median of the distribution, corresponding to the quantile $0.5$, is predicted correctly by the machine learning model, we would compare the value $0.5$ to the ratio of CDF values (again randomly chosen from the corresponding range of CDF values for discrete target values, as described in sec. \ref{sec:cdf_histo}) of observed sales records being lower/higher than $0.5$. In other words, in case of the median, 50 percent of the ex post observed target values should be observed below the median of the corresponding individual predicted PDF and 50 percent above.

In order to judge whether the overall shape of the predicted distributions is predicted correctly, we repeat this procedure for a range of quantiles, for example $q = 0.1, 0.2, \ldots, 0.9$. However, we are free to choose which quantiles to look at, and in specific situations it might be advisable to look at the tails of the distribution in more detail, to make sure that even relatively rare events are estimated correctly by the machine learning algorithm, and add more quantiles for comparison in the region between, say $q = 0.95$ and $q = 0.99$. In the following, we call this method {\em inverse quantile profile plot}. Profile plots are akin to scatter plots and described in more detail in appendix \ref{sec:profile}.

\fig \ref{fig:invquant_example} illustrates five different collections of inverse quantile profile plots (each collection comparing to 6 specified quantile values, namely $q = 0.1$, $q = 0.3$, $q = 0.5$, $q = 0.7$, $q = 0.9$, and $q = 0.97$), for separate sets of exemplary PDF estimations and observed data combinations. The dashed horizontal lines indicates the fraction we expect, i.e., the specified quantile value, if the predictions are correct. For example, for the median, the line at $q = 0.5$ indicates that  50 percent of all PDF prediction and observed data combinations in a given data set should fall above the line, and 50 percent should fall below the line. The observation of the number of samples, indicated with different markers, that do in fact fall above and below a particular line, then allows the evaluation of the accuracy of PDF estimations. In case that the PDFs are not estimated correctly, the fractions will deviate from their expected values and the corresponding profile plot allows to judge whether for example the tails of the predicted distribution describing rare events are particularly problematic.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/invquant_example}
\caption{\label{fig:invquant_example} Inverse quantile profile plots comparing all predicted PDFs to the corresponding observed events. In the leftmost two columns, the outcome is illustrated if the estimate of the variance is biased, the center and center-right columns illustrate cases for which the mean estimation is biased, and the far right column shows the expected behavior if all predictions are correct.}
\end{center}
\end{figure}

In the same way as the method of filling all observed CDF values of the individual PDF predictions in a histogram and compare to a uniform distribution (described in sec. \ref{sec:cdf_histo}), inverse quantile profile plots do not work on individual PDF predictions but require a statistical population. However, calculating the fractions of the inverse quantile profile plots globally, i.e., over all samples in the data set, might not reveal certain deficiencies for a subset, e.g., a specific store or product group in the example of retail demand forecasts. Therefore, we combine the approach described above with the method of profile plots, where the quantity on the $x$-axis of the inverse quantile profile plot can be any variable of the data set at hand.

In summary, this approach has two major benefits compared to the method discussed in sec. \ref{sec:cdf_histo}: First, by explicitly visualizing several different quantiles, the inverse quantile profile plot reveals which part of the predicted PDF, such as the tails of the probability distribution, are particularly problematic. Second, by showing the dependency from the (arbitrary) variable on the $x$-axis, the inverse quantile profile plot reveals deviations of the predicted PDF from the actuals for different parts, e.g. specific categories, of that variable. Two examples for this can be found in Figs. \ref{fig:invquant_mean} and \ref{fig:invquant_dayofweek} in the next section.

@@

The methods discussed so far allow a detailed qualitative evaluation of PDF predictions. However, in order to also quantify the quality of the PDF predictions, a metric is required that assesses the difference between the distributions. Therefore, we want to compare the histogram of CDF observations of the predicted PDFs with the expected uniform distribution and define an accuracy measure in the range between 0 and 1, which takes the value of 1 when both distributions agree perfectly. Several approaches that measure the difference between two probability distributions are suggested in the literature, such as the first Wasserstein distance \cite{olkin1982}, also known as earth mover distance (EMD), the Kullback-Leibler divergence \cite{kullback1951}, also known as relative entropy, or the Jensen-Shannon divergence \cite{dagan1997}, also known as information radius.

Compared to the other mentioned methods, the first Wasserstein distance, representing the symmetric distance between two probability distributions on a given metric space, is more sensitive to smaller deviations, because it exhibits a linear behavior around zero (reflecting perfect agreement). Therefore, we focus on the first Wasserstein distance as the measure of difference in the following. For our purposes here, it can be defined by:
\begin{equation}
\text{EMD}(P, Q) = \frac{\sum_{k=1}^N |F_P(x_k) - F_Q(x_k)|}{N},
\end{equation}
where $F_P(X)$ and $F_Q(X)$ are the CDFs of the two PDFs $P(X)$ and $Q(X)$, respectively, and $x_k$ denotes the average value of $X$ in bin $k$, with $X$ being divided in $N$ bins.

Since $0.5$ represents the maximum value of the first Wasserstein distance (its minimum being zero), we define an accuracy measure for our PDF predictions in the range $[0, 1]$ by:
\begin{equation}
\text{accuracy} = 1 - 2 \cdot \text{EMD}
\end{equation}

@@

\noindent
In this model focusing on the variance, we use the same set of features as for the mean model described above in sec. \ref{sec:example_mean}, except for dropping most of the two-dimensional combinations (only keeping \texttt{item\_id} - \texttt{store\_id}, \texttt{store\_id} - day of week, and \texttt{item\_id} - event type features) and adding the in-sample mean prediction as feature. An example for the resulting PDF and CDF predictions for item \texttt{FOODS\_3\_516} in store \texttt{TX\_3} on 2016-05-06, together with the corresponding observed sales value, is shown in \fig \ref{fig:pdf_example}.

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/pdf}
\includegraphics[width=4cm]{../figures/cdf}
\caption{\label{fig:pdf_example} Predicted PDF (left) and CDF (right) for a specific product-location-date combination. The dashed vertical line in the CDF plot represents the actually observed sales, corresponding to a CDF value just under $0.3$.}
\end{center}
\end{figure}

@@

Using the inverse quantile profile plots introduced earlier, the prediction quality of the full predicted PDF can be assessed in more detail. In \fig \ref{fig:invquant_dayofweek}, each column corresponds to a day of the week, using 0 for Mondays, 1 for Tuesdays, etc., and the considered quantiles are 0.1, 0.3, 0.5, 0.7, 0.9, and 0.97. Again, we compare the outcome of our negative binomial model to a simpler Poisson assumption. Aggregating over all stores, items, and sales dates (independently for each day of the week), we can see that for the negative binomial model each quantile is predicted relatively well, except for Sundays showing a tendency for overprediction of the mean (see \fig \ref{fig:invquant_example}). The Poisson model, on the other hand, shows a significant pattern of too narrow PDF predictions across all days of the week.

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/invquant_dayofweek_nbinom}
\includegraphics[width=4cm]{../figures/invquant_dayofweek_poisson}
\caption{\label{fig:invquant_dayofweek} Inverse quantile profile plot for the different days of week on the $x$-axis (from Monday to Sunday) aggregated over all product-location-day combinations in the test period, for our negative binomial model (left) and a simpler Poisson model for comparison (right).}
\end{center}
\end{figure}

\fig \ref{fig:invquant_mean} shows the inverse quantile profile plot using the predicted mean as the $x$-axis of the graph, meaning that the distributions are grouped such that the 15 columns correspond to mean predictions in 12 intervals with widths of 5 from 0.0 to 60.0 and 3 remaining intervals $[60.0, 70.0]$, $(70.0, 80.0]$, and $(80.0, 100.0]$ (with mean predictions higher than 100.0 included in the highest interval), while we aggregate over all locations, items, and sales dates. The considered quantiles are chosen to be the same as in \fig \ref{fig:invquant_dayofweek}. While the Poisson model again shows a significant pattern of too narrow PDF predictions across the full range of mean prediction values, we can also see several deviations from the expected uniform behavior for the negative binomial model. For mean predictions around $30$, the negative binomial model deviates significantly from the expected behavior, with the shape of the deviations pointing towards too broad PDF predictions (as also observed in the CDF histogram above). And for mean predictions higher than $80$, the pattern of the deviations indicate an underprediction of the mean parameter. The deviations at very low predicted mean values reflect the complications of slow-sellers described in the context of CDF histograms above.

@@

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/invquant_yhat_mean_nbinom}
\includegraphics[width=4cm]{../figures/invquant_yhat_mean_poisson}
\caption{\label{fig:invquant_mean} Inverse quantile profile plot for mean predictions on the $x$-axis aggregated over all product-location-day combinations in the test period, for our negative binomial model (left) and a simpler Poisson model for comparison (right).}
\end{center}
\end{figure}

The quantitative results for the CDF accuracy of our PDF predictions using the first Wasserstein distance as metric (as described in sec. \ref{sec:cdf_acc}) and calculated over all product-location-day combinations in the test period, can be found in \tab \ref{tab:cdf_acc}. As benchmark, we again compare against a Poisson model assumption using the same mean predictions. As expected from the qualitative findings in \fig \ref{fig:cdf_demand}, the negative binomial PDF predictions show a significant improvement over the simpler Poisson model.

\begin{table}[h!]
\begin{center}
\caption{Accuracy for negative binomial and Poisson PDF predictions, using the first Wasserstein distance as metric, calculated over all product-location-day combinations in the test period.}
\label{tab:cdf_acc}
\begin{tabular}{c|c|c}
 & \textbf{NBD} & \textbf{Poisson} \\
\hline
\textbf{EMD accuracy} & 0.967 & 0.850
\end{tabular}
\end{center}
\end{table}

@@

Once the full probability density functions are forecasted, their evaluation poses significant challenges: Common metrics, such as the mean absolute deviation or mean squared error, only take point estimates into account and are generally not suitable for the evaluation of predicted probability distributions. Furthermore, since the distributions are predicted for individual events, statistical techniques aiming at the comparison of distributions using many events cannot be used in general. Instead, novel techniques exploiting the probability integral transform, namely the histogram of observed CDF values of the predicted individual PDFs as well as inverse quantile profile plots, allow a detailed investigation into the behavior of the predictions. Finally, a quantitative assessment resulting in a single number can be obtained using metrics such as the earth mover distance in a comparison between the histogram of observed CDF values and the expected uniform distribution.
